Efficient simulation of individual-based population models

### 2024-04-01 article.qmd
proofread

Stochastic IBMs allow the modeling in continuous time of populations dynamics structured by age and/or characteristics. In the field of mathematical biology and ecology, a large community has used this formalism for the study of the evolution of structured populations (see e.g. [@FerTra09; @collet2013rigorous; @BanMel15; @costa2016stochastic; @billiard2016effect; @lavallee2019stochastic; @meleard2019birth; @calvez2020horizontal]), after the pioneer works [@FouMel04;@tran_2008;@meleard2009trait].

IBMs are also useful in demography and actuarial sciences, for the modeling of human populations dynamics (see e.g. [@Ben10; @Bou16; @karoui2021simulating]). They allow the modeling of heterogeneous and complex population dynamics, which can be used to compute demographic indicators or simulate the evolution of insurance portfolios in order to study the basis risk, compute cash flows for annuity products or pension schemes, or for a fine assessment of mortality models [@barrieu2012understanding]. There are other domains in which stochastic IBMs can be used, for example in epidemiology with stochastic compartmental models, neurosciences, cyber risk, or Agent-Based Models (ABMs) in economy and social sciences, which can be seen as IBMs.
Many mathematical results have been obtained in the literature cited above, for quantifying the limit behaviors of IBMs in long time or in large population. In particular, pathwise representations of IBMs have been introduced in [@FouMel04] (and extended to age-structured populations in [@tran_2008;@meleard2009trait]), as measure-valued pure jumps Markov processes, solutions of SDEs driven by Poisson measures. These pathwise representations are based on the *thinning* and projection of Poisson random measures defined on extended spaces. However, the simulation of large and interacting populations is often referred as computationally expensive.

@@

We opted to implement the algorithms of the `IBMPopSim` package using the `Rcpp` package, a tool facilitating the seamless integration of high-performance `C++` code into easily callable `R` functions [@JSSv040i08]. With just a few lines of `C++` code, `IBMPopSim` offers user-friendly R functions for defining IBMs. Once events and their associated intensities are specified, an automated procedure creates the model. This involves integrating the user's source code into the primary `C++` code using a template mechanism. Subsequently, `Rcpp` is invoked to compile the model and integrate it into the `R` session. Following this process, the model becomes callable with varying parameters, enabling the generation of diverse population evolution scenarios. Combined with the design of the simulation algorithms, the package structure yields very competitive simulation runtimes for IBMs, while staying user-friendly for `R` users. Several outputs function are also implemented in `IBMPopSim`. For instance the package allows the construction and visualization of age pyramids, as well as the construction of death and exposures table from the censored individual data, compatible with `R` packages concerned with mortality modelling, such as [@Rdemography] or [@Rstmomo]. Several examples are provided in the form of `R` vignettes on the [website](https://daphnegiorgi.github.io/IBMPopSim/), and in recent works of [@karoui2021simulating] and [@roget2022positive].

Designed for applications in social sciences, the `R` package `MicSim` [@Zin14] can be used for continuous time microsimulation. In continuous-time microsimulation, individual life-courses are usually specified by sequences of state transitions (events) and the time spans between these transitions. The state space is usually discrete and finite, which is no necessarily the case in `IBMPopSim`, where individuals can have continuous characteristics. But most importantly, microsimulation does not allow for interactions between individuals. Indeed, microsimulation produces separately the life courses of all individuals in the populations, based on the computation of the distribution functions of the waiting times in the distinct states of the state space, for each individual [@Zin14]. This can be slow in comparison to the simulation by thinning of event times occurring in the population, which is based on selecting event times among some competing proposed event times. Finally, `MicSim` simplifies the Mic-Core microsimulation tool implemented in Java [@zinn2009mic]. However, the implementation in `R` of simulation algorithms yields longer simulation run times than when using `Rcpp`. To the best of our knowledge, there are no other `R` packages currently available addressing the issue of IBMs efficient simulation.

@@

The number of individuals present in the population *before time* $t$ is obtained by taking $f\equiv 1$:
$$
    N_t =  \int_{\mathcal{I}}  Z_t(\mathrm{d}\tau^b, \mathrm{d}\tau^d, \mathrm{d}x) = \sum_{k=1}^{N_t} \boldsymbol{1}_{\mathcal{I}} (I_k).
$$
Note that $(N_t)_{t\geq 0}$ is an increasing process since dead/exited individuals are kept in the population $Z$. The number of alive individuals in the population at time $t$ is: 
$$
    N_t^a =  \int_{\mathcal{I}}  \mathsf{1}_{\{\tau^d > t \} }Z_t(\mathrm{d}\tau^b, \mathrm{d}\tau^d, \mathrm{d}x) = \sum_{k=1}^{N_t} \mathsf{1}_{\{\tau^d_k > t \} }.
$$ {#eq-Nta}
Another example is the number of alive individuals of age over $a$ is
$$
    N_t([a,+\infty)) :=  \int_{\mathcal{I}}  \boldsymbol{1}_{[a,+\infty)}(t-\tau^b)\mathsf{1}_{]t,\infty]}(\tau^d) Z_t(\mathrm{d}\tau^b, \mathrm{d}\tau^d, \mathrm{d}x) = \sum_{k=1}^{N_t}  \boldsymbol{1}_{\{ t -\tau_k^b \geq a \}}\mathsf{1}_{\{\tau^d_k \geq t \} }.
$$

@@

Event | Type | $\phi^e(t, I)$ | New individual
------|:----:|----------------|---------------
Birth | $b$ | $\delta_{(t, \infty,  X)}$ | $\tau^b =t, \; X \sim k^b(t,I,\mathrm{d}x)$ 
Entry | $en$ | $\delta_{(\tau^b, \infty,  X)}$ | $(\tau^b, X) \sim k^{en}(t,\mathrm{d}s, \mathrm{d}x)$ 
Death/Exit | $d$ | $\delta_{(\tau^b, t,x)} - \delta_{(\tau^b, \infty, x)}$ | $\tau^d = t$
Swap | $s$ | $\delta_{(\tau^b, \infty , X)} - \delta_{(\tau^b, \infty, x)}$ | $X \sim k^s(t,I,\mathrm{d}x)$
: Action in the population for a given event name {#tbl-event-action}

@@

- At time $T$, the population $Z_T$ contains all individuals who lived in the population before $T$, including dead/exited individuals. If there are no swap events, or entries,the population state $Z_t$ for any time $t\leq T$ can be obtained from $Z_T$. Indeed, if $Z_T = \sum_{k=1}^{N_T}  \delta_{I_k}$, then the population at time $t\leq T$ is simply composed of the individuals born before $t$: $$
    Z_t = \sum_{k=1}^{N_T} \boldsymbol{1}_{\{\tau^b_k \leq t \}} \delta_{I_k}.
$$
- In the presence of entries (open population), a characteristic $x$ can track the individuals' entry dates. Then, the previous equation can be easily modified in order to obtain the population $Z_t$ at time $t\leq T$ from $Z_T$.

@@

Once the different event types have been defined in the population model, the frequency at which each event occur in the population $e$ have to be specified.
Informally, the intensity $\Lambda^e_t(Z_t)$ at which an event $e$ can occur is defined by
$$
    \mathbb P\big( \text{event } e \text { occurs in the population }  Z_t  \in (t,t+\mathrm{d}t] | \mathcal{F}_t \big) \simeq  \Lambda^e_t (Z_t)\mathrm{d}t.
$$
For a more formal definition of stochastic intensities, we refer to [@bremaud1981point] or [@KaaElK20]. The form of the intensity function $(\Lambda^e_t (Z_t))$ determines the population simulation algorithm in `IBMPopSim`:

@@

::: {#thm-eq-Z} 
## Pathwise representation
Let $T\in {\mathbb{R}}^+$ and $\mathcal{J}= \mathbb N \times(\mathcal E \cup \mathcal{E}_W)$. 
Let $Q$ be a random Poisson measure on $\mathbb R^+ \times \mathcal{J}\times \mathbb{R}^+$, of intensity $\mathrm{d}t \delta_{\mathcal{J}}(\mathrm{d}k,\mathrm{d}e)  \mathbf{1}_{[0,\bar \lambda^e]} (\theta)\mathrm{d}\theta$, with $\delta_{\mathcal{J}}$ the counting measure on $\mathcal{J}$. 
Finally, let $Q^{\mathcal P}$ be a random Poisson measure on $\mathbb R^+ \times \mathcal{P}  \times \mathbb{R}^+$, of intensity $\mathrm{d}t \delta_{\cal P}(\mathrm{d}e)  \mathbf{1}_{[0,\bar \mu^e]} (\theta)\mathrm{d}\theta$, and $Z_0= \sum_{k=1}^{N_0} \delta_{I_k}$ an initial population.
Then, under @cnj-non-explosion
, there exists a unique
measure-valued population process $Z$, strong solution on the following
SDE driven by the Poisson measure $Q$: 
$$
    \begin{aligned}
    Z_t = Z_0 &+ \int_0^t \int_{ \mathcal{J}\times \mathbb R^+ }\phi^e (s , I_k)  \mathbf{1}_{\{k \leq N_{s^-}\} }\mathbf{1}_{\{\theta \leq \lambda_s^e(I_k, Z_{s^-})\}} Q (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ) \\
    &+   \int_0^t \int_{\mathcal{P} \times \mathbb R^+}  \phi^e(s, I_{s^-}) \mathbf{1}_{\{\theta \leq \mu^e(s) \}} Q^{\mathcal{P}} (\mathrm{d}s ,\mathrm{d}e,  \mathrm{d}\theta),  \qquad \forall  0 \leq t \leq T,
\end{aligned}
$$ {#eq-sde-eq-Z}
and where $I_{s^-}$ is an individual, chosen uniformly among alive individuals in the population $Z_{s^-}$.
:::

@@

By @prp-marked-poisson , $\bar{Q}$ is a Poisson random measure with mean measure
$$
    \bar \mu(\mathrm{d}t, \mathrm{d}\theta): = \bar \Lambda \mathrm{d}t
    \frac{\mathbf{1}_{[0, \bar \Lambda]}(\theta)}{\bar \Lambda} \mathrm{d}\theta= \mathrm{d}t  \mathbf{1}_{[0, \bar \Lambda]}(\theta) \mathrm{d}\theta.
$$
In particular, the average number of atoms $(\bar T_\ell, \bar \Theta_\ell)$ in $[0,t]\times [0,h]$ is
$$
    \mathbb{E}[Q([0,t]\times [0,h])]={\mathbb{E}}[\sum_{\ell} \boldsymbol{1}_{[0,t]\times [0,h]} (\bar T_\ell, \bar \Theta_{\ell})]  = \int_{(\mathbb{R}^+)^2}  \bar \mu(\mathrm{d}t, \mathrm{d}\theta)  = t (\bar \Lambda \wedge h).
$$
The thinning is based on the restriction property for Poisson measure: for a measurable set $\Delta\subset {\mathbb{R}}^+\times {\mathbb{R}}^+$, the restriction $Q^\Delta:= \boldsymbol{1}_{\Delta}\bar Q$ of $\bar Q$ to $\Delta$ (by taking only atoms in $\Delta$) is also a Poisson random measure of mean measure $\mu^{\Delta}(\mathrm{d}t, \mathrm{d}\theta)  = \boldsymbol{1}_{\Delta}(t,\theta) \bar \mu(\mathrm{d}t, \mathrm{d}\theta).$

In order to obtain an inhomogeneous Poisson measure of intensity $(\Lambda(t))$, the "good" choice of $\Delta$ is the hypograph of $\Lambda$:
$\Delta =\{ (t,\theta) \in [0,T]\times [0,\bar \Lambda] ; \; \theta \leq \Lambda(t)\}$ 
(see @fig-thinning). Then, 
$$
    Q^\Delta = \sum_{\ell \ge 1} \mathbf{1}_{\left\{\bar \Theta_\ell \le \Lambda(\bar T_\ell)\right\}} \delta_{(\bar T_\ell, \bar \Theta_\ell)},
$$ 
and since $\Lambda(t) \leq \bar \Lambda$, on $[0,T]$:
$$
\begin{aligned}
    \mu^{\Delta}(\mathrm{d}t, \mathrm{d}\theta) & = \boldsymbol{1}_{\{ \theta \leq \Lambda(t)\}}  \mathrm{d}t
\mathbf{1}_{[0, \bar \Lambda]}(\theta)\mathrm{d}\theta = \boldsymbol{1}_{\{\theta \leq \Lambda(t)\}} \mathrm{d}t \mathrm{d}\theta.
\end{aligned}
$$

@@

::: {#prp-thinning}
The counting process $N^\Lambda$, projection of $Q^{\Delta}$ on the time component and defined by,
$$
    N^{\Lambda}_t := Q^{\Delta}( [0,t] \times {\mathbb{R}}^+) = \int_0^t \int_{{\mathbb{R}}^+} \boldsymbol{1}_{\{\theta \leq \Lambda(s)\}} \bar Q(\mathrm{d}s, \mathrm{d}\theta) = \sum_{\ell \geq 1} \mathsf{1}_{\{ \bar T_\ell \leq t \}} \mathsf{1}_{\{\bar \Theta_\ell \leq  \Lambda(\bar T_\ell) \}},  \quad \forall t \in [0,T],
$$ {#eq-thinning-inhomogeneous-poisson}
is an inhomogeneous Poisson process on $[0,T]$ of intensity function $(\Lambda(t))_{t\in [0,T]}$. The thinning @eq-thinning-inhomogeneous-poisson is a pathwise representation of $N^\Lambda$ by *restriction and projection* of the Poisson measure $Q$ on $[0,T]$.
:::

@@

By @prp-marked-poisson and @prp-proj-poisson-measure, each measure $\bar Q_j (\mathrm{d}t, \mathrm{d}\theta) = \bar Q(\mathrm{d}t, \{j\}, \mathrm{d}\theta) = \sum_{\ell \geq 1} \mathsf{1}_{\{\bar J_\ell=j \}} \delta_{(\bar T_\ell, \bar \Theta_\ell)}$ is a marked Poisson measure of intensity
$$
    \bar \mu_j ( \mathrm{d}t  ,\mathrm{d}\theta) = \bar{\Lambda}p_j \mathrm{d}t  \frac{\mathsf{1}_{\{\theta \leq \bar \lambda_{j}\}}(\theta)}{\bar \lambda_{j}} \mathrm{d}\theta = \mathrm{d}t  \mathsf{1}_{\{\theta \leq \bar \lambda_{j}\}}(\theta) \mathrm{d}\theta.
$$
As a direct application of @prp-thinning , the inhomogeneous multivariate Poisson process is obtained by restriction of each measures $\bar Q_j$ to $\Delta_j = \{ (t, \theta) \in [0,T] \times [0,\bar \lambda_j] ;\; \theta \leq \lambda_j(t) \}$ and projection.

::: {#prp-thinning-vector}
The multivariate counting process $(N^j)_{j \in \mathcal{J}}$, defined for all $j \in \mathcal{J}$ and $t \in [0,T]$ by thinning and projection of $\bar Q$:
$$
    N^j_t  := \int_0^t \int_{ {\mathbb{R}}^+} \boldsymbol{1}_{\{\theta \leq \lambda_j(s)\}} \bar{Q}(\mathrm{d}s, \{j\}, \mathrm{d}\theta) = \sum_{\ell \geq 1} \mathsf{1}_{\{ \bar T_\ell \leq t \}}\mathsf{1}_{\{\bar J_\ell = j\}} \mathsf{1}_{\{\bar \Theta_\ell \leq  \lambda_j (\bar T_\ell ) \}},
$$
is an inhomogeneous Poisson process of intensity vector $(\lambda_j(t))_{j\in\mathcal{J}}$ on $[0,T]$.
:::

@@

The steps for simulating the first event in the population can be iterated in order to simulate the population. At the $k$th step, the same procedure is repeated to simulate the $k$th event, starting from a population $Z_{T_{k-1}}$ of size $N_{T_{k-1}}$.

@@

We now present the main algorithm implemented in the `popsim` function of the `IBMPopSim` package in the case where events arrive with individual intensities, but also with interactions (using randomization) and Poisson intensities. In this general case, $\bar \Lambda(n)$ is defined by @eq-def-bar-Lambda.

```pseudocode
#| label: alg-randomized
#| html-indent-size: "1.2em"
#| html-comment-delimiter: "//"
#| html-line-number: true
#| html-line-number-punc: ":"
#| html-no-end: false
#| pdf-placement: "htb!"
#| pdf-line-number: true

@@

### Individual

In the `C++` model which is automatically generated and compiled, an individual `I` is an object of an internal class containing some attributes (`birth_date`, `death_date` and the characteristics), and some methods including:

@@

### Events bounds

@@

@@

#### Entry events

The age of the new individual is determined by the `kernel_code`
argument in the `mk_event_poisson` function.

@@

#### Death and exit events

The baseline death intensity defined
in @eq-insurance-baseline and obtained with the package
`StMoMo` is stored in the variable `death_male`.

>>>>

Stochastic IBMs allow the modeling in continuous time of populations dynamics structured by age and/or characteristics. In the field of mathematical biology and ecology, a large community has used this formalism for the study of the evolution of structured populations (see e.g. @FerTra09; @collet2013rigorous; @BanMel15; @costa2016stochastic; @billiard2016effect; @lavallee2019stochastic; @meleard2019birth; @calvez2020horizontal), after the pioneer works [@FouMel04;@tran_2008;@meleard2009trait].

IBMs are also useful in demography and actuarial sciences, for the modeling of human populations dynamics (see e.g. @Ben10; @Bou16; @karoui2021simulating). They allow the modeling of heterogeneous and complex population dynamics, which can be used to compute demographic indicators or simulate the evolution of insurance portfolios in order to study the basis risk, compute cash flows for annuity products or pension schemes, or for a fine assessment of mortality models [@barrieu2012understanding]. There are other domains in which stochastic IBMs can be used, for example in epidemiology with stochastic compartmental models, neurosciences, cyber risk, or Agent-Based Models (ABMs) in economy and social sciences, which can be seen as IBMs.
Many mathematical results have been obtained in the literature cited above, for quantifying the limit behaviors of IBMs in long time or in large population. In particular, pathwise representations of IBMs have been introduced in @FouMel04 (and extended to age-structured populations in @tran_2008;@meleard2009trait), as measure-valued pure jumps Markov processes, solutions of SDEs driven by Poisson measures. These pathwise representations are based on the *thinning* and projection of Poisson random measures defined on extended spaces. However, the simulation of large and interacting populations is often referred as computationally expensive.

@@

We opted to implement the algorithms of the `IBMPopSim` package using the `Rcpp` package, a tool facilitating the seamless integration of high-performance `C++` code into easily callable `R` functions [@JSSv040i08]. With just a few lines of `C++` code, `IBMPopSim` offers user-friendly R functions for defining IBMs. Once events and their associated intensities are specified, an automated procedure creates the model. This involves integrating the user's source code into the primary `C++` code using a template mechanism. Subsequently, `Rcpp` is invoked to compile the model and integrate it into the `R` session. Following this process, the model becomes callable with varying parameters, enabling the generation of diverse population evolution scenarios. Combined with the design of the simulation algorithms, the package structure yields very competitive simulation runtimes for IBMs, while staying user-friendly for `R` users. Several outputs function are also implemented in `IBMPopSim`. For instance the package allows the construction and visualization of age pyramids, as well as the construction of death and exposures table from the censored individual data, compatible with `R` packages concerned with mortality modelling, such as @Rdemography or @Rstmomo. Several examples are provided in the form of `R` vignettes on the [website](https://daphnegiorgi.github.io/IBMPopSim/), and in recent works of @karoui2021simulating and @roget2022positive.

Designed for applications in social sciences, the `R` package `MicSim` [@Zin14] can be used for continuous time microsimulation. In continuous-time microsimulation, individual life-courses are usually specified by sequences of state transitions (events) and the time spans between these transitions. The state space is usually discrete and finite, which is no necessarily the case in `IBMPopSim`, where individuals can have continuous characteristics. But most importantly, microsimulation does not allow for interactions between individuals. Indeed, microsimulation produces separately the life courses of all individuals in the populations, based on the computation of the distribution functions of the waiting times in the distinct states of the state space, for each individual [@Zin14]. This can be slow in comparison to the simulation by thinning of event times occurring in the population, which is based on selecting event times among some competing proposed event times. Finally, `MicSim` simplifies the Mic-Core microsimulation tool implemented in Java [@zinn2009mic]. However, the implementation in `R` of simulation algorithms yields longer simulation run times than when using `Rcpp`. To the best of our knowledge, there are no other `R` packages currently available addressing the issue of stochastic IBMs efficient simulation.

@@

The number of individuals present in the population *before time* $t$ is obtained by taking $f\equiv 1$:
$$
    N_t =  \int_{\mathcal{I}}  Z_t(\mathrm{d}\tau^b, \mathrm{d}\tau^d, \mathrm{d}x) = \sum_{k=1}^{N_t} \mathbf{1}_{\mathcal{I}} (I_k).
$$
Note that $(N_t)_{t\geq 0}$ is an increasing process since dead/exited individuals are kept in the population $Z$. The number of alive individuals in the population at time $t$ is: 
$$
    N_t^a =  \int_{\mathcal{I}}  \mathbf{1}_{\{\tau^d > t \} }Z_t(\mathrm{d}\tau^b, \mathrm{d}\tau^d, \mathrm{d}x) = \sum_{k=1}^{N_t} \mathbf{1}_{\{\tau^d_k > t \} }.
$$ {#eq-Nta}
Another example is the number of alive individuals of age over $a$ is
$$
    N_t([a,+\infty)) :=  \int_{\mathcal{I}}  \mathbf{1}_{[a,+\infty)}(t-\tau^b)\mathbf{1}_{]t,\infty]}(\tau^d) Z_t(\mathrm{d}\tau^b, \mathrm{d}\tau^d, \mathrm{d}x) = \sum_{k=1}^{N_t}  \mathbf{1}_{\{ t -\tau_k^b \geq a \}}\mathbf{1}_{\{\tau^d_k \geq t \} }.
$$

@@

Event | Type | $\phi^e(t, I)$ | New individual
------|:----:|----------------|---------------
Birth | $b$ | $\delta_{(t, \infty,  X)}$ | $\tau^b =t, \; X \sim k^b(t,I,\mathrm{d}x)$ 
Entry | $en$ | $\delta_{(\tau^b, \infty,  X)}$ | $(\tau^b, X) \sim k^{en}(t,\mathrm{d}s, \mathrm{d}x)$ 
Death/Exit | $d$ | $\delta_{(\tau^b, t,x)} - \delta_{(\tau^b, \infty, x)}$ | $\tau^d = t$
Swap | $s$ | $\delta_{(\tau^b, \infty , X)} - \delta_{(\tau^b, \infty, x)}$ | $X \sim k^s(t,I,\mathrm{d}x)$
: Action in the population for a given event type {#tbl-event-action}

@@

- At time $T$, the population $Z_T$ contains all individuals who lived in the population before $T$, including dead/exited individuals. If there are no swap events, or entries,the population state $Z_t$ for any time $t\leq T$ can be obtained from $Z_T$. Indeed, if $Z_T = \sum_{k=1}^{N_T}  \delta_{I_k}$, then the population at time $t\leq T$ is simply composed of the individuals born before $t$: $$
    Z_t = \sum_{k=1}^{N_T} \mathbf{1}_{\{\tau^b_k \leq t \}} \delta_{I_k}.
$$
- In the presence of entries (open population), a characteristic $x$ can track the individuals' entry dates. Then, the previous equation can be easily modified in order to obtain the population $Z_t$ at time $t\leq T$ from $Z_T$.

@@

Once the different event types have been defined in the population model, the frequency at which each event occur in the population $e$ have to be specified.
Informally, the intensity $\Lambda^e_t(Z_t)$ at which an event $e$ can occur is defined by
$$
    \mathbb P\big( \text{event } e \text { occurs in the population }  Z_t  \in (t,t+\mathrm{d}t] | \mathcal{F}_t \big) \simeq  \Lambda^e_t (Z_t)\mathrm{d}t.
$$
For a more formal definition of stochastic intensities, we refer to @bremaud1981point or @KaaElK20. The form of the intensity function $(\Lambda^e_t (Z_t))$ determines the population simulation algorithm in `IBMPopSim`:

@@

::: {#thm-eq-Z} 
## Pathwise representation
Let $T\in {\mathbb{R}}^+$ and $\mathcal{J}= \mathbb N \times(\mathcal E \cup \mathcal{E}_W)$. 
Let $Q$ be a random Poisson measure on $\mathbb R^+ \times \mathcal{J}\times \mathbb{R}^+$, of intensity $\mathrm{d}t \delta_{\mathcal{J}}(\mathrm{d}k,\mathrm{d}e)  (\theta)\mathrm{d}\theta$, with $\delta_{\mathcal{J}}$ the counting measure on $\mathcal{J}$. 
Finally, let $Q^{\mathcal P}$ be a random Poisson measure on $\mathbb R^+ \times \mathcal{P}  \times \mathbb{R}^+$, of intensity $\mathrm{d}t \delta_{\cal P}(\mathrm{d}e)\mathrm{d}\theta$, and $Z_0= \sum_{k=1}^{N_0} \delta_{I_k}$ an initial population.
Then, under @cnj-non-explosion
, there exists a unique
measure-valued population process $Z$, strong solution on the following
SDE driven by the Poisson measure $Q$: 
$$
    \begin{aligned}
    Z_t = Z_0 &+ \int_0^t \int_{ \mathcal{J}\times \mathbb R^+ }\phi^e (s , I_k)  \mathbf{1}_{\{k \leq N_{s^-}\} }\mathbf{1}_{\{\theta \leq \lambda_s^e(I_k, Z_{s^-})\}} Q (\mathrm{d}s ,\mathrm{d}k , \mathrm{d}e, \mathrm{d}\theta ) \\
    &+   \int_0^t \int_{\mathcal{P} \times \mathbb R^+}  \phi^e(s, I_{s^-}) \mathbf{1}_{\{\theta \leq \mu^e(s) \}} Q^{\mathcal{P}} (\mathrm{d}s ,\mathrm{d}e,  \mathrm{d}\theta),  \qquad \forall  0 \leq t \leq T,
\end{aligned}
$$ {#eq-sde-eq-Z}
and where $I_{s^-}$ is an individual, chosen uniformly among alive individuals in the population $Z_{s^-}$.
:::

@@

By @prp-marked-poisson , $\bar{Q}$ is a Poisson random measure with mean measure
$$
    \bar \mu(\mathrm{d}t, \mathrm{d}\theta): = \bar \Lambda \mathrm{d}t
    \frac{\mathbf{1}_{[0, \bar \Lambda]}(\theta)}{\bar \Lambda} \mathrm{d}\theta= \mathrm{d}t  \mathbf{1}_{[0, \bar \Lambda]}(\theta) \mathrm{d}\theta.
$$
In particular, the average number of atoms $(\bar T_\ell, \bar \Theta_\ell)$ in $[0,t]\times [0,h]$ is
$$
    \mathbb{E}[Q([0,t]\times [0,h])]={\mathbb{E}}[\sum_{\ell} \mathbf{1}_{[0,t]\times [0,h]} (\bar T_\ell, \bar \Theta_{\ell})]  = \int_{(\mathbb{R}^+)^2}  \bar \mu(\mathrm{d}t, \mathrm{d}\theta)  = t (\bar \Lambda \wedge h).
$$
The thinning is based on the restriction property for Poisson measure: for a measurable set $\Delta\subset {\mathbb{R}}^+\times {\mathbb{R}}^+$, the restriction $Q^\Delta:= \mathbf{1}_{\Delta}\bar Q$ of $\bar Q$ to $\Delta$ (by taking only atoms in $\Delta$) is also a Poisson random measure of mean measure $\mu^{\Delta}(\mathrm{d}t, \mathrm{d}\theta)  = \mathbf{1}_{\Delta}(t,\theta) \bar \mu(\mathrm{d}t, \mathrm{d}\theta).$

In order to obtain an inhomogeneous Poisson measure of intensity $(\Lambda(t))$, the "good" choice of $\Delta$ is the hypograph of $\Lambda$:
$\Delta =\{ (t,\theta) \in [0,T]\times [0,\bar \Lambda] ; \; \theta \leq \Lambda(t)\}$ 
(see @fig-thinning). Then, 
$$
    Q^\Delta = \sum_{\ell \ge 1} \mathbf{1}_{\left\{\bar \Theta_\ell \le \Lambda(\bar T_\ell)\right\}} \delta_{(\bar T_\ell, \bar \Theta_\ell)},
$$ 
and since $\Lambda(t) \leq \bar \Lambda$, on $[0,T]$:
$$
\begin{aligned}
    \mu^{\Delta}(\mathrm{d}t, \mathrm{d}\theta) & = \mathbf{1}_{\{ \theta \leq \Lambda(t)\}}  \mathrm{d}t
\mathbf{1}_{[0, \bar \Lambda]}(\theta)\mathrm{d}\theta = \mathbf{1}_{\{\theta \leq \Lambda(t)\}} \mathrm{d}t \mathrm{d}\theta.
\end{aligned}
$$

@@

::: {#prp-thinning}
The counting process $N^\Lambda$, projection of $Q^{\Delta}$ on the time component and defined by,
$$
    N^{\Lambda}_t := Q^{\Delta}( [0,t] \times {\mathbb{R}}^+) = \int_0^t \int_{{\mathbb{R}}^+} \mathbf{1}_{\{\theta \leq \Lambda(s)\}} \bar Q(\mathrm{d}s, \mathrm{d}\theta) = \sum_{\ell \geq 1} \mathbf{1}_{\{ \bar T_\ell \leq t \}} \mathbf{1}_{\{\bar \Theta_\ell \leq  \Lambda(\bar T_\ell) \}},  \quad \forall t \in [0,T],
$$ {#eq-thinning-inhomogeneous-poisson}
is an inhomogeneous Poisson process on $[0,T]$ of intensity function $(\Lambda(t))_{t\in [0,T]}$. The thinning @eq-thinning-inhomogeneous-poisson is a pathwise representation of $N^\Lambda$ by *restriction and projection* of the Poisson measure $Q$ on $[0,T]$.
:::

@@

By @prp-marked-poisson and @prp-proj-poisson-measure, each measure $\bar Q_j (\mathrm{d}t, \mathrm{d}\theta) = \bar Q(\mathrm{d}t, \{j\}, \mathrm{d}\theta) = \sum_{\ell \geq 1} \mathbf{1}_{\{\bar J_\ell=j \}} \delta_{(\bar T_\ell, \bar \Theta_\ell)}$ is a marked Poisson measure of intensity
$$
    \bar \mu_j ( \mathrm{d}t  ,\mathrm{d}\theta) = \bar{\Lambda}p_j \mathrm{d}t  \frac{\mathbf{1}_{\{\theta \leq \bar \lambda_{j}\}}(\theta)}{\bar \lambda_{j}} \mathrm{d}\theta = \mathrm{d}t  \mathbf{1}_{\{\theta \leq \bar \lambda_{j}\}}(\theta) \mathrm{d}\theta.
$$
As a direct application of @prp-thinning , the inhomogeneous multivariate Poisson process is obtained by restriction of each measures $\bar Q_j$ to $\Delta_j = \{ (t, \theta) \in [0,T] \times [0,\bar \lambda_j] ;\; \theta \leq \lambda_j(t) \}$ and projection.

::: {#prp-thinning-vector}
The multivariate counting process $(N^j)_{j \in \mathcal{J}}$, defined for all $j \in \mathcal{J}$ and $t \in [0,T]$ by thinning and projection of $\bar Q$:
$$
    N^j_t  := \int_0^t \int_{ {\mathbb{R}}^+} \mathbf{1}_{\{\theta \leq \lambda_j(s)\}} \bar{Q}(\mathrm{d}s, \{j\}, \mathrm{d}\theta) = \sum_{\ell \geq 1} \mathbf{1}_{\{ \bar T_\ell \leq t \}}\mathbf{1}_{\{\bar J_\ell = j\}} \mathbf{1}_{\{\bar \Theta_\ell \leq  \lambda_j (\bar T_\ell ) \}},
$$
is an inhomogeneous Poisson process of intensity vector $(\lambda_j(t))_{j\in\mathcal{J}}$ on $[0,T]$.
:::

@@

The steps for simulating the first event in the population can be iterated in order to simulate the population. At the $k$th step, the same procedure is repeated to simulate the $k$th event, starting from a population $Z_{T_{k-1}}$ of size $N_{T_{k-1}}$. The algorithm is summarized in @alg-interaction.

@@

We now present the main @alg-randomized implemented in the `popsim` function of the `IBMPopSim` package in the case where events arrive with individual intensities, but also with interactions (using randomization) and Poisson intensities. In this general case, $\bar \Lambda(n)$ is defined by @eq-def-bar-Lambda.

```pseudocode
#| label: alg-randomized
#| html-indent-size: "1.2em"
#| html-comment-delimiter: "//"
#| html-line-number: true
#| html-line-number-punc: ":"
#| html-no-end: false
#| pdf-placement: "H"
#| pdf-line-number: true

@@

_Individual_ In the `C++` model which is automatically generated and compiled, an individual `I` is an object of an internal class containing some attributes (`birth_date`, `death_date` and the characteristics), and some methods including:

@@

_Events bounds_

@@

If there are no events with intensity of class `interaction`, then the
simulation can be parallelized easily by setting the optional parameter
`multithreading` (`FALSE` by default) to `TRUE`. By default, the number
of threads is the number of concurrent threads supported by the
available hardware implementation. The number of threads can be set
manually with the optional argument `num_threads`. By default, when the
proportion of dead individuals in the population exceeds $10\%$, dead
individuals are removed from the current population used in the
algorithm (see @rem-remvoval-dead). The user can modify this
ratio using the optional argument `clean_ratio`, or by removing dead
individuals from the population with a certain frequency, given by the
`clean_step` argument. Finally, the user can also define the seed of the
random number generator stored in the argument `seed`.

_Outputs and treatment of swap events_

The output of the `popsim` function is a list containing three elements:
a data frame `population` containing the output population $Z_T$ (or a
list of populations $(Z_{t_1}, \dots Z_{t_n})$ if `time` is a vector of
times), a numeric vector `logs` of variables related to the simulation
algorithm (including the simulation time and number of proposed/accepted
events), and the list `arguments` of the simulation inputs, including
the initial population, parameters and event bounds used for the
simulation.

When there are no swap events (individuals don't change of
characteristics), the evolution of the population over the period
$[0,T]$ is recorded in a single data frame `sim_out$population` where
each line contains the information of an individual who lived in the
population over the period $[0,T]$ (see @rem-popfinale).

When there are swap events (individuals can change of characteristics),
recording the dates of swap events and changes of characteristics
following each swap event and for each individual in the population is a
memory intensive and computationally costly process. To maintain
efficient simulations in the presence of swap events, the argument
`time` of `popsim` can be defined as a vector of dates
$(t_0,\dots, t_n)$. In this case, `popsim` returns in the object
`population` a list of $n$ populations representing the population at
time $t_1,\dots t_n$, simulated from the initial time $t_0$. For
$i=1\dots n$, the $i$th data frame is the population $Z_{t_i}$, i.e.
individuals who lived in the population during the period $[t_0,t_i]$,
with their characteristics at time $t_i$.

It is also possible to isolate the individuals' life course, by adding
an `id` column to the population, which can be done by setting `id=TRUE`
in the population construction, or by calling the `add_characteristic`
function to an existing population, in order to identify each individual
with a unique integer.

Base functions to study the simulation outputs are provided in the
package. For instance, the population age pyramid can computed at a give
time, as well as death and exposure tables. Several illustrations of the
outputs functions are given in the example @sec-insurance-portfolio and
@sec-example-interaction.

@@

**Entry events** The age of the new individual is determined by the `kernel_code`
argument in the `mk_event_poisson` function.

@@

**Death and exit events** The baseline death intensity defined
in @eq-insurance-baseline and obtained with the package
`StMoMo` is stored in the variable `death_male`.

```{r, results='hide'}
# StMoMo death rates
library('StMoMo')
library('reshape2')
EWStMoMoMale <- StMoMoData(EWdata_hmd, series = "male")
LC <- lc()
ages.fit <- 65:100
years.fit <- 1950:2016
LCfitMale <- fit(LC, data = EWStMoMoMale, ages.fit = ages.fit, years.fit = years.fit)
t <- 30
LCforecastMale <- forecast(LCfitMale, h = t)
d_k <- apply(LCforecastMale$rates, 2, function(x) stepfun(66:100, x))
breaks <- 1:29
death_male <- piecewise_xy(breaks,d_k)
```
